Improving the modelling of tempo and phoneme duration

### 2017-05-02 Paper/Paper_v2.tex
Fixed formatting errors and corrected some chapters in the paper. See Paper_v2 for changes.

\documentclass[11pt,a4paper]{scrbook}
\usepackage{geometry}	
\usepackage[backend=bibtex,style=numeric]{biblatex}
\usepackage{ngerman}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[style=numeric]{biblatex}
\bibliography{../PaperBib}
\usepackage{colortbl}	
\usepackage{xcolor}
\usepackage{soul}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[outdir=../Graphen/]{epstopdf}
\usepackage{textcomp}
\usepackage{floatrow}
\renewcommand{\figurename}{Fig.}
\usepackage{caption}
\setcapindent{0pt}
\usepackage{subcaption}
\usepackage{scrhack}
\renewcommand{\familydefault}{\sfdefault}
\definecolor{uhhred}{cmyk}{0,100,100,0}
\usepackage{tabularx}

@@

@@

\tableofcontents

\chapter{Introduction}
Some people speak faster, others speak slower, and all people speak sometimes faster or slower as they usually do. This ``speed`` at which one speaks is what we call ``speech rate``. The fact that we understand people at all natural speech rates is certainly largely due to the quality features of sounds (phonemes) in the given language. However, if we have a perfect audio recording of speech, and play it faster or slower, we immediately notice how naturalness deteriorates. Synthetic speech shows a similar behavior when changing the speaking rate. In both cases phoneme quality doesn't need to change. It is its duration that changes, along with some other suprasegmental features like pitch, phrasal accents, etc., and that do the trick.

@@

\subsection*{Scope}
Phoneme length adjustment is the aspect we wish to adress in this work. In particular, we want to find an algorithm for adjusting phoneme durations according to the variation of the speech rate. The resulting phoneme duration should be close to the actual duration that occurs when the speaking rate alters. This would be the same as playing an audio recording faster or slower.

There are several models for phoneme duration approximation that work pretty well for some languages, German inclusive. However, they consistently don't take the speech rate into consideration, their performance level being acquired rather within a specific speech rate. When this varies, model performance deteriorates, like van Santen states for his model \cite{Santen1994}. Moreover, all these models have an upper asymptote for their accuracy situated at less than 100\%. It has been suggested, that the influence of so-called ``macroscopic``or para-linguistic \cite{Santen1994} factors should be examined in order to move this asymptote further up. Such a factor is speech rate, and we took the challenge of integrating it into a speech model.

The purpose is therefore not to create another duration approximation model for TTS-systems, but to improve the algorithm of phoneme duration prediction by considering the tempo of speech. Table \ref{tab:Question} visualizes our challenge: we want to put all available/visible information from the table in relation so that we can replace the question marks with values that are close to the real ones. Consequently, the modified speech unit would sound more natural at a faster or a slower speech rate as when being modified with a linear function.

\begin{table}[htbp]
\caption {The challenge} 
\centering
\begin{tabular}{|L{2.5cm}||*{4}{C{.8cm}|}|*{4}{C{.8cm}|}}
	\hline
			& \multicolumn{4}{c||}{ \textit{genau}} & \multicolumn{4}{c|}{ \textit{genau}} \\
	\hline
Phonemes		& g & @ & n & aU & g & @ & n & aU \\
	\hline
Durations (sec) & 0.05 & 0.07 & 0.05 & 0.11 & ?  & ?  & ?  & ? \\
	\hline
Speech rate $dur/\#phon$ & \multicolumn{4}{c||}{0.07} & \multicolumn{4}{c|}{0.11} \\
	\hline
\end{tabular}
\caption{The first column shows our inputs: Duration of each phoneme in German word ``genau``, given a speech rate of 0.07 sec, or a word duration of 0.28 sec. The second column visualizes our challenge: How long is the duration of each phoneme of the given word, at a speech rate of 0.11 sec, equivalent of a word length of 0.42 sec?}
\label{tab:Question} 
\end{table}

Hoequist and Kohler observed already back in 1986 that the change in speaking tempo does not produce a linear change of the acoustic segments of an utterance (i. e. phonemes) \cite{Hoequist1986}. Indeed, if one simply ``stretches`` all phonemes in the same manner to accommodate a new speech rate, the result is rather disappointing, as one can see in Fig. \ref{fig:linear_approx}.
\begin{figure}[htbp]
	\centering
	\includegraphics[scale=0.7]{../Graphen/Linear_approx.eps}
	\caption[Linear phoneme duration approximation]{This is a linear solution to the challenge presented in table \ref{tab:Question}.}
	\label{fig:linear_approx}
\end{figure}

Across the paper we use the German SAMPA notations when referring to phonemes.

\subsection*{Outline}
We start our challenge by defining our methods in chapter 2. In the following chapter we examine the available resources so that we can proceed with the actual task in chapter 4, where we analyze the German phonemes. We dedicate chapter 5 to the challenge of finding an adequate speech rate definition for our database, and which can serve our purpose. Chapter 6 finally deals with phoneme duration prediction models and the challenge of finding a solution to our above presented issue.

\chapter{Means and Methods}

@@

We used machine learning to solve a task involving both classification and numeric prediction, and selected from its repertoire two model tree methods: M5P and REPTree. The reasons we chose are: trees are easy to interpret, they can perform both classification and numeric prediction tasks, we could easily generate them using Weka, and they are the closest to classification and regression (CART) trees, also used in the literature \cite{Brinckmann_2003} for phoneme duration prediction, so we have data to compare with.

\subsection{Decision trees}
Decision trees divide the attribute space into well defined clusters, identified by a class name or a numeric value. In the first case we are dealing with a classification tree, in the second one with a regression tree. If one needs to combine the two methods, he/she may use a classification and regression tree, called CART, or a model tree. The advantage of trees is that new instances are easily assigned to a class, or a numeric value, or a linear model which leads to a numeric value, and they are easy to unterstand.

@@

\section{Performance Evaluation}
Models produced by machine learning may have different performance levels, which may be evaluated in several ways. As they assess different features of the models, and treat outliers differently, we didn't want to rely only on one of them. Furthermore, we wanted to be able to compare the results with results presented in the literature. After studying the common evaluation methods used for other phoneme prediction models, we chose the following three: 
\begin{enumerate}
	\item \textbf{RMSE} or \textbf{root mean squared error} measures the differences between the numeric values predicted by a model and those actually observed: 
$RMSE = \sqrt{\frac{\sum(predicted-actual)^2}{n}}$. It is a very popular method for evaluating errors of a model, but it is very sensible to outliers.
	\item \textbf{MAE} or \textbf{mean absolute error} is similar to RMSE, but penalises outliers less. It is computed using the average of the absolute errors: $MAE = \frac{\sum\left|predicted-actual\right|}{n}$
	\item The \textbf{Pearson correlation coefficient}, a popular measure for continuous data, is a number between -1 and 1 that quantifies the correlation between two variables, each having a separate value set. We can interpret this as how much the variable movement through its value set corresponds to the movement of the other variable in its own value set. Values close to 1 mean that greater values of the one variable imply greater values for the second variable, negative values mean the two variables move in opposite directions, while values close to 0 mean there is no correlation between the two variables. However, one must keep in mind that correlation does not imply causality, so the presence of correlation between two variables does not mean that one of them has any influence on the other one. If we compare red Ferrari cars to blue Renault cars, one may find a correlation between car color red and car speed. However, this does not imply that red cars are generally faster.
\end{enumerate}

@@

In order to extract the needed information from our database, and put it in a form ready to use with Weka, we used self-made code written in Python 3.6. This is an easy to learn programming language with many features dedicated to statistical analysis and even machine learning. Most of the figures in this paper were produced in python as well.

\chapter{Corpora}
Language related databases are called corpora, and for our purpose we needed one containing segmented and annotated German speech as a set of text files, so-called ``textual data``. Our minimum requirement for such a corpus was that it contains speech segmented at phoneme level, allowing us to calculate the duration of each phoneme occurence. The unit in which this duration is expressed is not relevant. As phoneme duration may be influenced by several factors, and we did not intent to do any further processing of the audio files, we also needed that this textual database contains enough information to identify following attributes:

@@

There are three corpora which we had at our disposal: (1) The Kiel Corpus of Read (PhonDat) and of Spontaneous Speech, (2) Verbmobil, and (3) spoken Wikipedia. The first one contains two equal sized corpora of > 4 hours each: KCoRS, which is a collection of read speech, and KCoSS, which is a collection of spontaneous speech, both being manually segmented, a good indicator for good segmentation quality. However, as Pfitzinger \cite{Pfitzinger2002} also showed in his reassessment of the PhonDat-II corpus, for nowadays research the Kiel Corpus is rather small. Spoken Wikipedia is a new project comprising of a much larger collection of read speech. Unfortunately, the phoneme-level segmentation was not ready on time to evaluate in our project.

@@

\chapter{German phoneme system}
The phoneme inventory used in Verbmobil includes 52 classes defined according to the German SAMPA. However, only 49 of these occur in the examined corpus. The missing ones are \texttt{/2/} as in \textit{Ökonom}, and the nasals \texttt{/E\~/} and \texttt{/O\~/} as in the loanwords \textit{Teint, Saison}. Using the classical phonetic conventions, we can group these phonemes using a tree structure. The first branching splits them into vowels, consonants and a glottal stop, although the latter one is sometimes treated as a consonant \cite{Kohler1995}. Vowels represent 24\% of the total phoneme inventory analyzed. Further relevant groupings are discussed in subsections ``Vowels`` and ``Consonants``.

Pauses may be caused by and/or be quantitatively modified both by linguistic (e.g. phonological) and non-linguistic factors, like technical, psychological, dialog dynamics, etc. Consequently pause length variation needs to be treated separately from phoneme duration variation, as a not negligible large set of other factors are to be considered for this, such as dialogue dynamics, hesitations, technical issues of the recording, which extends beyond the purpose of this work.

@@

One can say that vowels generally tend to get proportionally longer when the words themselves are spoken slower. We investigated this assumption by comparing the vowel duration inside words containing a specific number of phonemes, as well as the variation of the word time proportion occupied by them in such words. The results confirmed this hypothesis, as you may see in Fig. \ref{fig:w_dur_vs_pho} and \ref{fig:boxplot_prop}. This also means that it is a good idea to look for the correlation between vowels and speech rate. We obtained an overall correlation coefficient of $r = 0.72$ for vowels against speech rate. Inside the vowel group, we could identify the primary stressed vowels as the actual correlation owners, with a group correlation value of $r = 0.75$, while the unstressed and secondary stressed ones showed rather modest correlation values of $r = 0.56$, respectively $r = 0.47$.

\begin{figure}[htbp]
	\noindent\makebox[\textwidth]{\includegraphics[width=\textwidth, clip]{../Graphen/Stats_for_genau.eps}}
	\centering
	\vspace{-10mm}
	\caption[Word duration of ``genau`` vs. its phoneme durations]{We ordered ascendingly the 706 occurences of the word ``genau`` based on word duration and ploted recorded durations of the component phonemes against it. The picture suggests clearly that the diphthong \texttt{/aU/} occupies a rather fixed steak of the total word duration, while the other phonemes show no adjustment to the increasing word total duration. }
	\label{fig:w_dur_vs_pho}
\end{figure}

\begin{figure}[htbp]
	\includegraphics[width=.55\linewidth]{../Graphen/box3_oyauaiaab.eps}
	\centering
	\caption[Proportions of diphthong in 3-phoneme-words]{The time slice occupied by vowels in words of specific length  vary relatively little, indicating a possible linear correlation between phoneme duration and the variation of word duration. The y-axis measures time in seconds.}
	\label{fig:boxplot_prop}
\end{figure}

@@

Table \ref{tab:vowel_stats} below presents the main duration information regarding the above mentioned classes that consolidates the discussion presented above.
\begin{table}[htbp]
\centering
%\vspace{10mm}
\begin{tabular}{|l|c|c|c|c|c|c|c|}

\hline
	 & Min value & Max value & Mean & St. Dev. & Q25 & Q75 & Count\\
\hline
\hline
Diphthongs  	& .04  & .71 & .12 & .06 & .08 & .15 & 8495\\
	\hline
Long vowels   & .04 & 2.3 &  .11 & .08 & .06 & .13 & 30425\\
	\hline
Schwa  &  .03 & 1.02 & .08 & .07 & .04 & .09 & 21637 \\
	\hline
Short vowels & .03 & .81 & .07 & .04 & .04 & .08 & 42570 \\
	\hline
\end{tabular}
\caption{Numeric information on the duration of the main vowel classes. The unit of the values presented in all columns except ``Count`` is second.} 
\label{tab:vowel_stats}
\end{table}

@@

\begin{figure}[htbp]
	\includegraphics[width=.55\linewidth]{../Graphen/vow_art.eps}
	\centering
	\caption[Vowel square]{Vowels are produced by the air flow in an open vowel tract, and they differ based on the modification we add to the shape of this cavity. The only way for humans to produce such modifications are by moving the tongue and/or the lips. These changes are usually represented in the so-called vowel square, which may be further simplified to a vowel triangle. Here you can see how to place a vowel square containing the main vowels in the mouth cavity. }
	\label{fig:vow_square}
\end{figure}

\section{Glottal stop}
The glottal stop occurs 13 348 times and only word initially in our database, marked with \texttt{/Q/}. We decided to treat it separately because of its particular usage in German: it may occur only at the beginning of a word or a word stem and in front of a vowel. Some phonetic manuals don't event consider it a phoneme \cite{Ternes2012}. Furthermore, its acoustical properties make it look like a (filled) pause in speech, which represents a challenge for the transcription. As our files were segmented automatically, there is place for doubt about the accuracy of the collected data for \texttt{/Q/}.

However, unlike pauses we did consider it  for analysis, as it is has an important communicative function as delimiter for words and morphemes. It is in fact one of the important clues which help hearer understand speech \cite{Ternes2012}.

@@

\section{Consonants}
In general, we can say that consonants present different degrees of elasticity, which are, however, generally lower than the ones of vowels.

Based on the correlation between phoneme duration and speech rate we created a subgroup of consonants containing the nasals and \texttt{/x/, /C/, /h/} with a relatively good correlation value $\geq 0.7$, and considered plosives to be a subgroup because of their consistently low correlation $\leq 0.42$.

@@

Same as vowels, the consonant branch of the tree may be split further down into smaller categories. We chose to include them in our phoneme overview, as some of these categories proved to be relevant for the modelling of phoneme durations as well. Consequently, level 2 split shows for for the consonant node following categories:
\begin{itemize}
	\item Plosives, which are generally short, due to their articulation characteristics. Their production involves two separate phases: closure, when the vocal tract is completely closed, and release, when the accumulated air flow is let out as in a burst. Considering this, it is straightforward, that the first phase may be longer, limited only by the human physiology, while the second one is technically extremely short. Some speech corpora, like PhonDat, label the two phases differently. Verbmobil doesn't, so we had to take them as a whole. We keep these considerations in mind, when referring to the special case of \texttt{/t/} that we found in our corpus.
	\item Fricatives, which show a duration variation almost comparable to that of vowels. This fact can also be explained by their production process: the vocal tract is not completely closed, but the air flow is forced out through a narrowing at some point in the vocal tract, so that a friction sound is produced.
	\item Nasals also show an elasticity comparable to that of vowels. Again this is explainable through their production: the oral passage is blocked, while lowering the velum, so that the air can escape through the nose. Nasals may also constitute the nucleus of a syllable, when the vowel gets reduced. /m/, /n/, /N/.
	\item Liquids, also called approximants, are produced by partial closure of the mouth produced by the thongue. German knows two liquids \texttt{/l/, /r/}.   One relevant feature is that they have the greatest freedom to occur in consonant clusters, that is a group of consonants belonging to the same syllable part. This also means that they must be particularly short, as we identified vowels as the greater steak holders in word duration. Another fact we considered, is the \texttt{/r/} is relatively seldom in German, being often reduced to a schwa, and when it is articulated, then it is generally extremely short, which durations comparable to the ones of the glottal stop.
\end{itemize}

@@

\hline
	 & Min value & Max value & Mean & St. Dev. & Q25 & Q75 & Count\\
\hline
\hline
Plosives  	& .03  & 1.50 & .06 & .05 & .03 & .07 & 45477\\
	\hline
Fricatives   & .03 & 1.89 &  .06 & .05 & .03 & .08 & 51667\\
	\hline
Nasals  &  .03 & 1.46 & .08 & .08 & .04 & .09 & 39936 \\
	\hline
Liquids /l/ & .03 & 1.51 & .06 & .06 & .03 & .07 & 8019 \\
	\hline
Liquids /r/ & .03 & .46 & .05 & .03 &  .03 & .06 & 5453 \\
	\hline
\end{tabular}
\caption{} 
\label{tab:cons_stats}
\end{table}

\chapter{Speech rate}
Speech rate is a measurement of the ``tempo`` of speech, that is how quickly someone speaks. This may be seen as a discrete/qualitative measure, with discrete values as \textit{slow}, \textit{normal}, and \textit{fast} or as a continuous one, with continuous numerical values.  We opted for the continuous version, because it is quantifyable and so can be a useful input into a regression, see Section 2.1.1.

Expressed mathematically, speech rate can be defined similarly to speed in spatial displacement, which would be distance covered by time. In the case of speech, distance is not counted in the spatial domain but in relevant linguistic units. Other than in the case of distance, the length of the speech units is not standardized (like kilometers, miles, etc.). Moreover it's size is also expressed in units of time: phoneme \texttt{/a/} has the length of x seconds, miliseconds, etc.

>>>>

\documentclass[11pt,a4paper]{scrbook}
\usepackage{geometry}	
%\usepackage[backend=bibtex,style=numeric]{biblatex}
%\usepackage{ngerman}
%\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[style=numeric]{biblatex}
\bibliography{../PaperBib2}
\usepackage{colortbl}	
\usepackage{xcolor}
%\usepackage{soul}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[outdir=../Graphen/]{epstopdf}
\usepackage{textcomp}
%\usepackage{floatrow}
\renewcommand{\figurename}{Fig.}
\usepackage[skip=0pt]{caption}
\captionsetup{format=hang}
\setcapindent{0pt}
\usepackage{subcaption}
\usepackage{scrhack}
%\renewcommand{\familydefault}{\sfdefault}
\definecolor{uhhred}{cmyk}{0,100,100,0}
\usepackage{tabularx}

@@

\usepackage{hyperref}
\hypersetup{linktocpage}

@@

\tableofcontents\thispagestyle{empty}

\mainmatter

\chapter{Introduction}
Some people speak faster, others speak slower, and all people speak sometimes faster and sometimes slower than they usually do. The tempo at which one speaks is what we call ``speech rate``. The fact that we understand people at all natural speech rates is certainly largely due to the quality features of sounds (phonemes) in the given language. However, if we have a perfect audio recording of speech, and play it faster or slower, we immediately notice how naturalness deteriorates. Synthetic speech shows a similar behavior when changing tempo. In both cases phoneme quality doesn't need to change. It is its duration that changes, along with some other suprasegmental features like pitch, and phrasal accents that do the trick.

@@

\subsection*{Scope}
Phoneme length adjustment is the aspect we wish to adress in this work. In particular, we want to find an algorithm for adjusting phoneme durations to the variation of the speech rate. The resulting phoneme duration should be close to the actual duration that occurs when speaking rate alters. This would be the same as playing an audio recording faster or slower.

There are several models for phoneme duration approximation that work pretty well for some languages, German inclusive. However, they consistently don't take the speech rate into consideration, their performance level being acquired rather at a specific speech rate. When this varies, model performance deteriorates, like van Santen states for his model \cite{Santen1994}. Moreover, all of these models have an upper asymptote for their accuracy situated at less than 100\%. It has been suggested, that the influence of so-called ``macroscopic`` or para-linguistic \cite{Santen1994} factors should be examined in order to move this asymptote further up. Such a factor is speech rate, and we took the challenge of integrating it into a speech model.

The purpose is therefore not to create another duration approximation model for TTS-systems, but to improve the algorithm of phoneme duration prediction by considering speech tempo. We argue therefore, that speech tempo produces a non-linear alteration of the phoneme durations.

Table \ref{tab:Question} visualizes our challenge: we want to put all available, i.e. visible information from the table in relation so that we can replace the question marks with values that are close to the real ones. Consequently, the modified speech unit sounds more natural when tempo varies.

\begin{table}[htbp]
\centering
\begin{tabular}{|L{2.5cm}||*{4}{C{.8cm}|}|*{4}{C{.8cm}|}}
	\hline
			& \multicolumn{4}{c||}{ \textit{genau}} & \multicolumn{4}{c|}{ \textit{genau}} \\
	\hline
Phonemes		& g & @ & n & aU & g & @ & n & aU \\
	\hline
Durations (sec) & 0.05 & 0.07 & 0.05 & 0.11 & ?  & ?  & ?  & ? \\
	\hline
Speech rate $dur/\#phon$ & \multicolumn{4}{c||}{0.07} & \multicolumn{4}{c|}{0.11} \\
	\hline
\end{tabular}
\vspace{5mm}
\caption{Our inputs: duration of each phoneme in German word ``genau`` at a speech rate of 0.07 sec. corresponding to a word duration of 0.28 sec, and a new speech rate, 0.11 sec, equivalent to a word duration of 0.42 sec, for which we have the challenge of finding appropriate phoneme durations.}
\label{tab:Question}
\end{table}

Hoequist and Kohler observed already back in 1986 that the change in speaking tempo does not produce a linear change of the acoustic segments of an utterance (i. e. phonemes) \cite{Hoequist1986}. Indeed, if one simply ``stretches`` all phonemes in the same manner to accommodate a new speech rate, the result is rather disappointing, as one can see in Fig. \ref{fig:linear_approx}.

Across the paper we use the German SAMPA notations when referring to phonemes.
\begin{figure}[htbp]
	\centering
	\includegraphics[trim=23mm 23mm 0mm 0mm, scale=0.7]{../Graphen/Linear_approx.eps}
	\vspace{4mm}
	\caption[Linear phoneme duration approximation]{This is a linear solution to the challenge presented in Table \ref{tab:Question}.}
	\label{fig:linear_approx}	
\end{figure}

\subsection*{Outline}
We start our challenge by defining our methods in Chapter \ref{chap_2}. In the following chapter we examine the available resources so that we can proceed with the actual task in Chapter \ref{chap_phonemes}, where we analyze the German phonemes. We dedicate Chapter \ref{chap_SR} to the challenge of finding an adequate speech rate definition for our database, and which can serve our purpose. Chapter \ref{chap_6:models} finally deals with phoneme duration prediction models and the challenge of finding a solution to our above presented issue.

\chapter{Means and Methods}
\label{chap_2}

@@

We used machine learning to solve a task involving both classification and numeric prediction, and selected from its repertoire two model tree methods: M5P and REPTree. We chose these trees because: trees are easy to interpret, they can perform both classification and numeric prediction tasks, we could easily generate them using Weka, and they are the closest to classification and regression (CART) trees, also used in the literature \cite{Brinckmann_2003} for phoneme duration prediction, so we have data to compare with.

\subsection{Decision trees}
\label{dec_trees}
Decision trees divide the attribute space into well defined clusters, identified by a class name or a numeric value. In the first case we are dealing with a classification tree, in the second one with a regression tree. If one needs to combine the two methods, he/she may use a classification and regression tree, called CART, or a model tree. The advantage of trees is that new instances are easily assigned to a class, or a numeric value, or a linear model which leads to a numeric value, and they are easy to unterstand.

@@

\section{Performance Evaluation}
Models produced by machine learning may have different performance levels, which may be evaluated in several ways. As they assess different features of the models, and treat outliers differently, we didn't want to rely only on one of them. Furthermore, we wanted to be able to compare the results with results presented in the literature. After studying the common evaluation methods used for other phoneme prediction models, we chose the following three: 
\begin{enumerate}
	\item \textbf{RMSE} or \textbf{root mean squared error} measures the differences between the numeric values predicted by a model and those actually observed: 
$RMSE = \sqrt{\frac{\sum(predicted-actual)^2}{n}}$. It is a very popular method for evaluating errors of a model, but it is very sensible to outliers.
	\item \textbf{MAE} or \textbf{mean absolute error} is similar to RMSE, but penalises outliers less. It is computed using the average of the absolute errors: $MAE = \frac{\sum\left|predicted-actual\right|}{n}$
	\item The \textbf{Pearson correlation coefficient}, a popular measure for continuous data, is a number between -1 and 1 that quantifies the correlation between two variables, each having a separate value set. We can interpret this as how much the variable movement through its value set corresponds to the movement of the other variable in its own value set. Values close to 1 mean that greater values of the one variable imply greater values for the second variable, negative values mean the two variables move in opposite directions, while values close to 0 mean there is no correlation between the two variables. However, one must keep in mind that correlation does not imply causality, so the presence of correlation between two variables does not mean that one of them has any influence on the other one. If we compare red Ferrari cars to blue Renault cars, one may find a correlation between car color red and car speed. However, this does not imply that red cars are generally faster.
\end{enumerate}
Throughout this work we consistently mean the Pearson's correlation coefficient whenever we speak about correlation.

@@

In order to extract the needed information from our database, and put it in a form ready to use with Weka, we used self-made code written in \textbf{Python} 3.6. This is an easy to learn programming language with many features dedicated to statistical analysis and even machine learning. Most of the figures in this paper were produced in python as well.

\chapter{Corpora}
\label{chap_3}
Language related databases are called corpora, and for our purpose we needed one containing segmented and annotated German speech as a set of text files, so-called ``textual data``. Our minimum requirement for such a corpus was that it contains speech segmented at phoneme level, allowing us to calculate the duration of each phoneme occurence. The unit in which this duration is expressed is not relevant. As phoneme duration may be influenced by several factors, and we did not intent to do any further processing of the audio files, we also needed that this textual database contains enough information to identify following attributes:

@@

There are three corpora which we had at our disposal: (1) The Kiel Corpus of Read (\textit{PhonDat} or \textit{KCoRS}) and of Spontaneous Speech (\textit{KCoSS}), (2) Verbmobil, and (3) spoken Wikipedia. The first one contains two equal sized corpora of > 4 hours each: KCoRS, which is a collection of read speech, and KCoSS, which is a collection of spontaneous speech, both being manually segmented, a good indicator for good segmentation quality. However, as Pfitzinger \cite{Pfitzinger2002} also showed in his reassessment of the PhonDat-II corpus, for nowadays research the Kiel Corpus is rather small. Spoken Wikipedia is a new project comprising of a much larger collection of read speech. Unfortunately, the phoneme-level segmentation was not ready on time to evaluate in our project, so we chose Verbmobil.

@@

\chapter{German phoneme system}
\label{chap_phonemes}
The phoneme inventory used in Verbmobil includes 52 classes defined according to the German SAMPA. However, only 49 of these occur in the examined corpus. The missing ones are \texttt{/2/} as in \textit{Ökonom}, and the nasals \texttt{/E$\sim$/} and \texttt{/O$\sim$/} as in the loanwords \textit{Teint, Saison}. Using the classical phonetic conventions, we can group these phonemes using a tree structure. The first branching splits them into vowels, consonants and a glottal stop, although the latter one is sometimes treated as a consonant \cite{Kohler1995}. Vowels represent 24\% of the total phoneme inventory analyzed. Further relevant groupings are discussed in subsections ``Vowels`` and ``Consonants``.

Pauses may be caused by and/or be quantitatively modified both by linguistic (e.g. phonological) and non-linguistic factors, like technical, psychological, and dialog dynamics. Consequently pause length variation needs to be treated separately from phoneme duration variation, as a not negligible large set of other factors are to be considered for this, such as dialogue dynamics, hesitations, or technical issues of the recording which all extend beyond the purpose of this work.

@@

One can say that vowels generally tend to get proportionally longer when the words themselves are spoken slower. We investigated this assumption by comparing the vowel duration inside words containing a specific number of phonemes, as well as the variation of the word time proportion occupied by them in such words. The results confirmed this hypothesis. We illustrate this in Figure \ref{fig:w_dur_vs_pho} on the word ``genau``, where you can see that the length of the diphthong \texttt{/aU/} changes proportionally to the word duration, while the other phonemes seem to be unaffected by word duration. This leads to the idea that at least some phonemes tend to occupy a specific steak (proportion) of the total word duration in words with a specific composition accounting for a linear correlation with word duration in such cases. Figure \ref{fig:boxplot_prop} shows the proportion distribution of diphthongs, the longest long vowel \texttt{/a:/}, its short counterpart \texttt{/a/}, and a plosive consonant \texttt{/b/} in words containing a total of three phonemes. As expected, the diphthongs occupy the greater steak, while the consonant the smaller one. This also means that it is a good idea to look for the correlation between vowels and speech rate. We obtained an overall correlation coefficient of $r = 0.72$ for vowels against speech rate. Inside the vowel group, we could identify the primary stressed vowels as the actual correlation owners, with a group correlation value of $r = 0.75$, while the unstressed and secondary stressed ones showed rather modest correlation values of $r = 0.56$, respectively $r = 0.47$.

\begin{figure}[htbp]
	\noindent\makebox[\textwidth]{\includegraphics[trim=30mm 30mm 30mm 10mm, clip, width=\textwidth]{../Graphen/Stats_for_genau.eps}}
	\centering
	%\vspace{mm}
	\caption[Word duration of ``genau`` vs. its phoneme durations]{We ordered ascendingly the 706 occurences of the word ``genau`` based on word duration and ploted recorded durations of the component phonemes against it. The picture suggests clearly that the diphthong \texttt{/aU/} occupies a rather fixed steak of the total word duration, while the other phonemes show no adjustment to the increasing word total duration. }
	\label{fig:w_dur_vs_pho}
\end{figure}

@@

\begin{figure}[htbp]
	\includegraphics[trim=15mm 0mm 15mm 0mm, width=.55\linewidth]{../Graphen/box3_oyauaiaab.eps}
	\centering
	\caption[Proportions of diphthong in 3-phoneme-words]{The time slice occupied by vowels in words of specific length  vary relatively little, indicating a possible linear correlation between their duration and the variation of word duration. Here you can see the difference in steak occupied by diphthongs next to the longest monophthongs \texttt{/a:/}, \texttt{/a/}, and a consonant \texttt{/b/}. The y-axis measures time in seconds.}
	\label{fig:boxplot_prop}
\end{figure}

Table \ref{tab:vowel_stats} below presents the main duration information regarding the above mentioned vowel classes that consolidates the discussion presented above.
\begin{table}[htbp]
\centering
%\vspace{10mm}
\begin{tabular}{|l|c|c|c|c|c|c|c|}

\hline
	 & Min value & Max value & Mean & St. Dev. & Q25 & Q75 & Count\\
\hline
\hline
Diphthongs  	& .04  & .71 & .12 & .06 & .08 & .15 & 8495\\
	\hline
Long vowels   & .04 & 2.3 &  .11 & .08 & .06 & .13 & 30425\\
	\hline
Schwa  &  .03 & 1.02 & .08 & .07 & .04 & .09 & 21637 \\
	\hline
Short vowels & .03 & .81 & .07 & .04 & .04 & .08 & 42570 \\
	\hline
\end{tabular}
\vspace{5mm}
\caption{Numeric information on the duration of the main vowel classes. The unit of the values presented in all columns except ``Count`` is second.} 
\label{tab:vowel_stats}
\end{table}

@@

\begin{figure}[htbp]
	\includegraphics[width=.5\linewidth]{../Graphen/vow_art.eps}
	\centering
\vspace{5mm}
	\caption[Vowel square]{Vowels are produced by the air flow in an open vowel tract, and they differ based on the modification we add to the shape of this cavity. The only way for humans to produce such modifications are by moving the tongue and/or the lips. These changes are usually represented in the so-called vowel square, which may be further simplified to a vowel triangle. Here you can see how to place a vowel square containing the main vowels in the mouth cavity and how alone schwa lies in the middle of this square - for the production of schwa we don't need to act on the air flow. }
	\label{fig:vow_square}
\end{figure}

\section{Glottal stop}
\label{Q}
The glottal stop occurs 13 348 times and only word initially in our database, marked with \texttt{/Q/}. We decided to treat it separately because of its particular usage in German: it may occur only at the beginning of a word or a word stem and in front of a vowel. Some phonetic manuals don't event consider it a phoneme \cite{Ternes2012}. Furthermore, its acoustical properties make it look like a (filled) pause in speech, which represents a challenge for the transcription. As our files were segmented automatically, there is place for doubt about the accuracy of the collected data for \texttt{/Q/}.

However, unlike pauses we did consider it  for analysis, as it has an important communicative function as delimiter for words and morphemes. It is in fact one of the important clues which help hearer understand speech \cite{Ternes2012}.

@@

\section{Consonants}
\label{cons}
In general, we can say that consonants present different degrees of elasticity, which are, however, generally lower than the ones of vowels.

Based on the correlation between phoneme duration and speech rate we created a subgroup of consonants containing the nasals and \texttt{/x/, /C/, /h/} with a relatively good correlation value $\geq 0.7$, and considered plosives to be an independent subgroup because of their consistently low correlation $\leq 0.42$.

@@

Same as vowels, the consonant branch of the tree may be split further down into smaller categories. We chose to include them in our phoneme overview, as some of these categories proved to be relevant for the modelling of phoneme durations as well. Consequently, level 2 split of the consonant node shows following categories:
\begin{itemize}
	\item Plosives, which are generally short, due to their articulation characteristics. Their production involves two separate phases: closure, when the vocal tract is completely closed, and release, when the accumulated air flow is let out as in a burst. Considering this, it is straightforward, that the first phase may be longer, limited only by the human physiology, while the second one is technically extremely short. Some speech corpora, like PhonDat, label the two phases differently. Verbmobil doesn't, so we had to take them as a whole. We keep these considerations in mind, when referring to the special case of \texttt{/t/} that we found in our corpus.
	\item Fricatives, which show a duration variation almost comparable to that of vowels. This fact can also be explained by their production process: the vocal tract is not completely closed, but the air flow is forced out through a narrowing at some point in the vocal tract, so that a friction sound is produced.
	\item Nasals also show an elasticity comparable to that of vowels. Again this is explainable through their production: the oral passage is blocked, while lowering the velum, so that the air can escape through the nose. Nasals may also constitute the nucleus of a syllable, when the vowel gets reduced. German has three nasal consonants: /m/, /n/, /N/.
	\item Liquids, also called approximants, are produced by partial closure of the mouth produced by the thongue. German knows two liquids \texttt{/l/, /r/}.   One relevant feature is that they have the greatest freedom to occur in consonant clusters, that is a group of consonants belonging to the same syllable part. This also means that they must be particularly short, as we identified vowels to be the greater steak holders in word duration. Another fact we considered, is that \texttt{/r/} is relatively seldom in German, being often reduced to a schwa, and when it is articulated, then it is generally extremely short, which durations comparable to the ones of the glottal stop.
\end{itemize}

@@

\hline
	 & Min value & Max value & Mean & St. Dev. & Q25 & Q75 & Count\\
\hline
\hline
Plosives  	& .03  & 1.50 & .06 & .05 & .03 & .07 & 45477\\
	\hline
Fricatives   & .03 & 1.89 &  .06 & .05 & .03 & .08 & 51667\\
	\hline
Nasals  &  .03 & 1.46 & .08 & .08 & .04 & .09 & 39936 \\
	\hline
Liquids /l/ & .03 & 1.51 & .06 & .06 & .03 & .07 & 8019 \\
	\hline
Liquids /r/ & .03 & .46 & .05 & .03 &  .03 & .06 & 5453 \\
	\hline
\end{tabular}
\vspace{4mm}
\caption{} 
\label{tab:cons_stats}
\end{table}

\chapter{Speech rate}
\label{chap_SR}
Speech rate is a measurement of the tempo of speech, that is, how quickly someone speaks. This may be seen as a discrete (qualitative) measure, with discrete values such as \textit{slow}, \textit{normal}, and \textit{fast} or as a continuous one, with continuous numerical values. We opted for the continuous version, because it is quantifiable and so can be a useful input into a regression, see Section \ref{dec_trees} on decision trees.
